Nonlinear Tidal Dissipation in Binary White Dwarfs

### 2019-03-03 proposal/FINESST/Su_FINESST_2019.tex
notes + repropose
final proposal

\section{Background and Proposed Research}\label{s:1}

@@

Previous works have found that the dissipation of IGW can generate significantly
more energy than thermal radiation from the isolated WD surface and is thus a
major contributor to the WD energy budget\cite{fullerII,fullerIV}. However,
these works parameterized the wave breaking process in an ad hoc manner. The
details of dissipation, namely the location and spatial extent of the wave
breaking, affect the observable outcome: dissipation near the surface of the WD
can be efficiently radiated away and simply brightens the WD, while dissipation
deep in the WD envelope causes an energy buildup that results in energetic
flares\cite{tidal_novae}. Works in other fields based on numerical simulations
show that strongly nonlinear wave breaking behaves differently than predictions
based in linear and weakly nonlinear theory\cite{winters1994,barker_ogilvie}.
Such fully nonlinear numerical simulations have not been performed for WDs.

@@

Characterizing the location and spatial extent of tidal dissipation in WD
binaries will require numerical simulation to capture the turbulent cascade to
small scales that causes wave breaking. \textbf{I propose to study the dynamical
effects of tidal dissipation via nonlinear internal gravity wave (IGW) breaking
in binary WDs.} There are three specific goals of my proposed research:
\begin{itemize}
    \item Characterize the location, spatial extent, and other properties of
        wave breaking in realistic WD models via direct numerical simulation.
        The location and spatial extent of wave breaking will furnish a simple
        yet effective parameterization of tidal dissipation in a range of WD
        models. \autoref{s:2} describes the steps required to acheive this goal.

    \item Predict signatures of tidal dissipation over a wide range of possible
        WD systems. In particular, I will study the impact of tidal heating
        on the luminosity of WDs in binaries and explore the possibility of
        producing of observable flares. \autoref{s:3} details my plan to predict
        observational manifestations of tidal heating.

    \item Compute modified GW templates for LISA that account for changes in the
        phase evolution of the orbit due to tidal dissipation. \autoref{s:4}
        elaborates on how I will perform this computation.
\end{itemize}

\section{Nonlinear Tidal Dissipation}\label{s:2}

@@

IGW breaking has been studied in atmospheric sciences. The wave
breaking process proceeds as follows: Initially, as the IGW reaches nonlinear
amplitudes, it breaks down via the parametric subharmonic instability and
transfers energy and angular momentum from the wave to the mean flow of the
fluid\cite{drazin}. After the mean flow velocity reaches the
horizontal phase velocity of the IGW, a critical layer forms. Analytical
calculations show that the IGW is nearly completely absorbed at this critical
layer in the linear approximation and endows the atmosphere with a mean
horizontal flow\cite{booker_bretherton,hazel}. However, when this mean flow
absorption was numerically studied including full nonlinear interactions, new
phenomena not described by the linear theory (reflection off the critical layer
and sharpening of the mean flow) were observed\cite{jones_num,winters1994}; this
nonlinear behavior significantly affected the evolution of the atmosphere over
time. \emph{This highlights the importance of numerical simulation in capturing
the wave breaking process.}

To gain insight into the tidal dissipation process, I began by adapting the
spectral hydrodynamics code Dedalus\cite{dedalus} to study IGW breaking in a 2D
isothermal, stratified atmosphere. A spectral code like Dedalus is ideal for
simulating complex hydrodynamical phenomena, as spectral methods have no
inherent numerical viscosity and so can more accurately resolve the nonlinear
cascade to small length scales in wave breaking.

Working with Dr.\ Daniel Lecoanet (a post-doc at Princeton University, and one
of the authors of the Dedalus code) and my advisor, Prof.\ Dong Lai, I have
simulated the nonlinear evolution of an upward-propagating IGW wavetrain excited
at the bottom of the atmosphere. My simulations show the waves breaking and
depositing horizontal momentum in the fluid, causing the fluid to acquire an
average horizontal flow, consistent with previous studies\cite{fullerII}. I have
derived simple formulae for the location and spatial extent of the dissipation
zone where the IGW is absorbed by the fluid. More interestingly, my simulation
reveals a partial reflection of the IGW at the critical layer\cite{me}, a
phenomenon not considered in the current astrophysical literature but consistent
with the aforementioned results\cite{winters1994}. A sample simulation is
presented in \autoref{fig:nl_fluxes}. I am preparing these results for
publication\cite{me}.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{nl_fluxes.png}
    \caption{The evolution of the average horizontal flow velocity of the fluid
    $U_0$ (in units of $c_{ph,x}$ the horizontal phase velocity of the IGW) and
    horizontal momentum flux contained in the IGW $S_{px}(z)$ (in units of the
    total excited flux $S_0$) for one of my simulations. Different lines
    correspond to different times in the simulation. Times are measured in units
    of $N^{-1}$, the inverse of the Brunt-V\"ais\"al\"a frequency (or the
    buoyancy period), and heights in units of $H$ (the scale height of the
    density stratification). IGWs are excited at $z = 2H$ and propagate to
    higher $z$ before undergoing wave breaking. The sharp decrease in $S_{px}$
    is the location of the critical layer where the IGW breaks and drives $U_0$
    towards $c_{ph, x}$; it moves to lower $z$ over time. The apparent decrease
    in $S_{px}$ at later times indicates wave reflection off the critical
    layer.}\label{fig:nl_fluxes}
\end{figure}

@@

It is clear that full numerical modeling of IGW breaking is necessary to fully
capture tidal dissipation. The first aim of my proposal is to \textbf{extend my
preliminary results concerning IGW breaking in stratified atmospheres to
characterize the dynamics of nonlinear IGW breaking in realistic WD models.} Via
numerical simulation, I will develop models for tidal dissipation that can be
used to study long-term WD evolution under tidal heating without expensive
hydrodynamical simulations. I will continue my work in the following stages:
\begin{itemize}
    \item I will perform simulations examining the validity of my results in
        spherical geometries to capture tidal effects in WD binaries. I will
        continue to use Dedalus, which supports spherical coordinates. Although
        3D simulations are necessarily more complex than my 3D simulations of
        stratified atmospheres, the underlying dynamics of IGW breaking are the
        same.

@@

\section{Tidal Heating and Binary White Dwarf Evolution}\label{s:3}

@@

In~\cite{tidal_novae} (hereafter FL), the authors used MESA\cite{MESA} to study
the production of tidal novae in binary WDs. A simple two-zone parameterization
where the tidal heat is deposited throughout the outer zone was used to model
tidal dissipation. It was found that cool WDs in sufficiently compact binaries
(orbital period $\lesssim 15$ minutes) may incur a thermonuclear detonation of
the hydrogen envelope. WD binaries with such short orbital periods have been
observed, e.g.\ SDSS J065133+284423 has a period of 12.75 minutes\cite{12min}.
FL fitted the observed properties of SDSS J065133+284423 to their tidal heating
models and finds evidence for tidal heating of the secondary WD in SDSS
J065133+284423.

@@

With the tidal dissipation models I will develop, I will be able to perform a
binary WD evolution study similar to that in FL but with a realistic tidal
dissipation profile (instead of their parameterized model). The second aim of my
proposal is to \textbf{use my tidal dissipation profiles to simulate a binary WD
undergoing tidal heating and make comparisons to observational data.} The
location of the tidal heating is a key ingredient in determining whether
the deposited tidal energy can be efficiently transfered away or whether the WD
experiences sudden detonations. As such, an improved understanding of the
properties of tidal dissipation is important to proper forecasting of binary
WDs' thermodynamic evolution. Moreover, while only a few sufficiently compact WD
binaries were available at the time of writing of FL, \emph{Gaia} data releases
2 and 3 will provide many more compact WD binaries to examine for observational
manifestations of tidal heating.

@@

\section{Tidal Dissipation and LISA}\label{s:4}

@@

The final aim of my proposal is \textbf{to use my tidal dissipation model to
compute WD binary GW waveforms including tidal dissipation for use in the LISA
detection pipeline}. This aim is much less computationally expensive than it
appears: LISA-band WD binaries can be well described using the Newtonian
dynamics of two co-orbiting point masses\cite{DWD_pointmass}. The resultant GW
emission can then be accurately computed using the weak gravity quadrupole
approximation (see e.g.~\cite{peters,lsst_wd}). Under these two approximations,
the GW waveform can be computed analytically without resorting to numerical
relativity simulations at all. Thus, I will compute GW waveforms accounting for
the additional phase evolution due to tidal dissipation. I will publish my
corrected waveforms for use by LISA and the GW community.

@@

\end{document}

>>>>

\section{Introduction and Goals of Proposed Research}\label{s:1}

@@

Previous works have found that the dissipation of IGW can generate significantly
more energy than thermal radiation from the isolated WD surface and is thus a
major contributor to the WD energy budget\cite{fullerII,fullerIV}. However,
these works parameterized the wave breaking process in an ad hoc manner. The
details of dissipation, namely the location and spatial extent of the wave
breaking, affect the observable outcome: dissipation near the surface of the WD
can be efficiently radiated away and brightens the WD, while dissipation
deep in the WD envelope causes an energy buildup that results in energetic
flares\cite{tidal_novae}. Works in other fields based on numerical simulations
show that strongly nonlinear wave breaking behaves differently than predictions
based on linear and weakly nonlinear theory\cite{winters1994,barker_ogilvie}.
Such fully nonlinear numerical simulations have not been performed for WDs.

@@

Characterizing the location and spatial extent of tidal dissipation in WD
binaries will require numerical simulation to capture the turbulent cascade to
small scales that causes wave breaking. \textbf{I propose to study the dynamical
effects of tidal dissipation via nonlinear internal gravity wave (IGW) breaking
in binary WDs.} There are three specific objectives of my proposed research:
\begin{itemize}
    \item Objective A\@: Characterize the location, spatial extent, and other
        properties of wave breaking in realistic WD models via direct numerical
        simulation. The location and spatial extent of wave breaking will
        furnish a simple yet effective parameterization of tidal dissipation in
        a range of WD models. \autoref{s:2} describes the steps required to
        acheive this goal.

    \item Objective B\@: Predict signatures of tidal dissipation over a wide
        range of possible WD systems. In particular, I will study the impact of
        tidal heating on the luminosity of WDs in binaries and explore the
        possibility of producing observable flares. \autoref{s:3} details my
        plan to predict observational manifestations of tidal heating.

    \item Objective C\@: Compute modified GW templates for LISA that account for
        changes in the phase evolution of the orbit due to tidal dissipation.
        \autoref{s:4} elaborates on how I will perform this computation.
\end{itemize}

\section{Objective A\@: Nonlinear Tidal Dissipation}\label{s:2}

@@

IGW breaking has been studied in atmospheric sciences. The wave breaking process
proceeds as follows: Initially, as the IGW reaches nonlinear amplitudes, it
breaks down via the parametric instability and transfers energy and angular
momentum from the wave to the mean flow of the fluid\cite{drazin}. After the
mean flow velocity reaches the horizontal phase velocity of the IGW, a critical
layer forms. Analytical calculations show that the IGW is nearly completely
absorbed at this critical layer in the linear approximation and endows the
atmosphere with a mean horizontal flow\cite{booker_bretherton,hazel}. However,
when this mean flow absorption was numerically studied including full nonlinear
interactions, new phenomena not described by the linear theory (reflection off
the critical layer and sharpening of the mean flow) were
observed\cite{jones_num,winters1994}; this nonlinear behavior significantly
affected the evolution of the atmosphere over time. \emph{This highlights the
importance of numerical simulation in capturing the wave breaking process.}

To gain insight into the tidal dissipation process, I have already begun to
adapt the spectral hydrodynamics code Dedalus\cite{dedalus} to study IGW
breaking in a 2D isothermal, stratified atmosphere. A spectral code like Dedalus
is ideal for simulating complex hydrodynamical phenomena, as spectral methods
have no inherent numerical viscosity and so can more accurately resolve the
nonlinear cascade to small length scales in wave breaking.

Working with Dr.\ Daniel Lecoanet (a post-doc at Princeton University, and one
of the authors of the Dedalus code) and my advisor, Prof.\ Dong Lai, I have
simulated the nonlinear evolution of an upward-propagating IGW wavetrain excited
at the bottom of the atmosphere (see \autoref{fig:nl_fluxes} for an example). My
simulations show that waves break and deposit horizontal momentum in the fluid,
causing the fluid to acquire an average horizontal flow, consistent with
previous studies\cite{fullerII}. I have derived simple formulae for the location
and spatial extent of the dissipation zone where the IGW is absorbed by the
fluid. More interestingly, my simulation reveals a partial reflection of the IGW
at the critical layer\cite{me}, a phenomenon not considered in the current
astrophysical literature but consistent with the aforementioned
results\cite{winters1994}. I am preparing these results for
publication\cite{me}.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.8\textwidth]{nl_fluxes.png}
    \caption{A sample numerical simulation of the evolution of the nonlinear
    internal gravity wave (IGW) propagating upwards and breaking in a stratified
    isothermal atmosphere. The upper panel shows the evolution of the average
    horizontal flow velocity of the fluid $U_0$ (in units of $c_{ph,x}$ the
    horizontal phase velocity of the IGW), and the lower panel shows the
    horizontal momentum flux contained in the IGW $S_{px}(z)$ (in units of the
    total excited flux $S_0$). Different lines correspond to different times in
    the simulation. Times are measured in units of $N^{-1}$, the inverse of the
    Brunt-V\"ais\"al\"a frequency (or the buoyancy period), and heights in units
    of $H$ (the scale height of the density stratification). IGWs are excited at
    $z = 2H$ and propagate to higher $z$ before undergoing wave breaking. The
    sharp decrease in $S_{px}$ corresponds to the location of the critical layer
    where the IGW breaks and drives $U_0$ towards $c_{ph, x}$; it moves to lower
    $z$ over time. The apparent decrease in $S_{px}$ at later times indicates
    wave reflection off the critical layer.}\label{fig:nl_fluxes}
\end{figure}

@@

It is clear that full numerical modeling of IGW breaking is necessary to fully
capture tidal dissipation. The first aim of my proposal is to \textbf{extend my
preliminary results concerning IGW breaking in stratified atmospheres to
characterize the dynamics of nonlinear IGW breaking in realistic WD models.} Via
numerical simulation, I will develop models for tidal dissipation that can be
used to study long-term WD evolution under tidal heating. I will continue my
work in the following stages:
\begin{itemize}
    \item I will perform simulations examining the validity of my results in
        spherical geometries to capture tidal effects in WD binaries. I will
        continue to use Dedalus, which supports spherical coordinates. Although
        3D simulations are necessarily more complex than my 2D simulations of
        stratified atmospheres (see above), the underlying dynamics of IGW
        breaking are the same.

@@

\subsection{Feasibility and Availability of Resources}

In my preliminary 2D simulations, I use $256$ horizontal and $1024$ vertical
spectral modes. These simulations take about $2$ days to run on $32$ threads using
Dedalus\cite{dedalus} using adaptive timestepping. Based on these preliminary
results, I intend to perform 3D simulations at resolution $128 \times 128 \times
512$. Including some optimizations I am still developing, I anticipate these
simulations will take about 2 weeks each when performed using 64 threads.

The computational resources available to me at Cornell include a cluster with
104 cores/208 threads as well as multiple 32 core/64 thread machines. There are
sufficient to perform many such 3D simulations within the timespan of the
proposal.

\section{Objective B\@: Tidal Heating and Binary White Dwarf Evolution
}\label{s:3}

@@

In~\cite{tidal_novae} (hereafter FL), the authors used MESA\cite{MESA} to study
the production of tidal novae in binary WDs. A simple two-zone parameterization
where the tidal heat is deposited throughout the outer zone was used to model
tidal dissipation. It was found that cool WDs in sufficiently compact binaries
(orbital period $\lesssim 15$ minutes) may incur a thermonuclear detonation of
the hydrogen envelope. WD binaries with such short orbital periods have been
observed, e.g.\ SDSS J0651 has a period of 12.75 minutes\cite{12min}. FL fitted
the observed properties of J0651 to their tidal heating models and finds
evidence for tidal heating of the secondary WD in this system.

@@

With the tidal dissipation models I will develop in Objective A (see
\autoref{s:2}), I will be able to perform a binary WD evolution study similar to
that in FL but with a realistic tidal dissipation profile (instead of their
parameterized model). The second aim of my proposal is to \textbf{use my tidal
dissipation profiles to simulate a binary WD undergoing tidal heating and make
comparisons to observational data.} The location of the tidal heating is a key
ingredient in determining whether the deposited tidal energy can be efficiently
transfered away or whether the WD experiences sudden detonations. As such, an
improved understanding of the properties of tidal dissipation is important to
proper forecasting of binary WDs' thermodynamic evolution. Moreover, while only
a few sufficiently compact WD binaries were available at the time of writing of
FL, \emph{Gaia} data releases 2 and 3 will provide many more compact WD binaries
to examine for observational manifestations of tidal heating.

@@

As MESA is a 1D stellar evolution code, I will easily have sufficient
computational resources to carry out this investigation.

\section{Objective C\@: Tidal Dissipation and LISA}\label{s:4}

@@

The final aim of my proposal is \textbf{to use my tidal dissipation model to
compute WD binary GW waveforms including tidal dissipation for use in the LISA
detection pipeline}. This aim is much less computationally expensive than it
appears: LISA-band WD binaries can be well described using leading-order
post-Newtonian dynamics and including the effect of tidal
deformations.\cite{DWD_pointmass}. The resultant GW emission can then be
accurately computed using the weak gravity quadrupole approximation (see
e.g.~\cite{peters,lsst_wd}). Under these two approximations, the GW waveform can
be computed analytically without resorting to numerical relativity simulations
at all. Thus, I will compute GW waveforms accounting for the additional phase
evolution due to tidal dissipation. I will publish my corrected waveforms for
use by LISA and the GW community.

@@

\end{document}
% mesa school?
% finish FINESST
% e disp? plot

% Kushnir?
% Bouss?? maybe
